curatedMetagenomicData

### 2019-10-08 curatedMetagenomic_NS.Rmd
updated

# the relab=FALSE is essential as it Absolute Raw Count Data
pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)

#For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object. curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:

@@

# These are useful is filtering. 
#sample_variables(pseq.subset)
#get_variable(pseq.subset, sample_variables(pseq.subset)[4])

@@

p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon",
 add = "boxplot", fill = "mnd", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
  print(p1)
#ggsave("nation_violin_shannon.png")

@@

>>>>

# the relab=FALSE is essential as it Absolute Raw Count Data
pseq = ExpressionSet2phyloseq(eset,relab=FALSE ,simplify = TRUE, phylogenetictree = TRUE)
orig_pseq <- pseq
#For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object. curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:

@@

# These are useful is filtering. 
#sample_variables(pseq.subset)
#get_variable(pseq.subset, sample_variables(pseq.subset)[13])

@@

p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon",
 add = "boxplot", fill = "mnd", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
  print(p1)

  ggsave("nation_violin_shannon.png")
#ggsave("nation_violin_shannon.png")

@@

## Nationality Violin Plot
```{r nationality violin plot}

# Alpha diversity is the species richness of a place
# High alpha diversity = high no. of species

#pseq.meta$nationality <- factor(pseq.meta$country, levels = c("Scandinavia","EasternEurope","CentralEurope", "SouthEurope", "UKIE", "US"))

# Violin plot
# the number in the palettes must equal the number of groups in nationlaity.pairs
#p1 <- ggviolin(pseq.meta, x = "nationality", y = "Shannon",
# add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
#   print(p1)
# ggsave("nation_violin_shannon.png")

# p2 <- ggviolin(pseq.meta, x = "nationality", y = "InverseSimpson",
#  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
# print(p2)
# ggsave("nation_violin_simpson.png")

# p3 <- ggviolin(pseq.meta, x = "nationality", y = "fisher",
#  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("") + ylab("Fisher")
# print(p3)
# ggsave("nation_violin_fisher.png")

# plot_grid(p1, p2, p3, align = "h" , ncol = 2, nrow = 2)
# ggsave("nation_violin.png")

```{r}

# 
pseq <- pseq.subset
# This isnt working. 
p <- plot_richness(pseq, x="country", measures=c( "Shannon"))

ggsave("alpha_nationality.png")

## MND Violin Plot
```{r mnd violin plot}

# Add the diversity table to metadata
pseq.meta$Shannon <- tab$shannon
#pseq.meta$InverseSimpson <- tab$inverse_simpson

# Compare differences in Shannon index between mnd group of the study subjects
pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")

# Create a list of pairwise comaprisons
mnd <- levels(pseq.meta$mnd) # get the variables

# Make a pairwise list that we want to compare.
mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))

# Shannon
# the number in the palettes must equal the number of groups in bmi.pairs
p1 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))
p1 <- p1 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND Prevalence") + theme(legend.position="none")
print(p1)
ggsave("mnd_violin__shannon_stats.png")

# Inverse Simpson
#p2 <- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
#p2 <- p2 + stat_compare_means(comparisons = mnd.pairs) 
#print(p2)
#ggsave("mnd_violin_simpson.png")

## BETA DIVERSITY
Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
WE want to: quantify community divergence within a given sample set
Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.

```{r Phylum NMDS}

png("Phylum_NMDS.png")
pseq.ord <- ordinate(pseq, "NMDS", "bray")

p1 = plot_ordination(pseq, pseq.ord, type = "taxa", color = "mnd", title = "Taxa by Phylum")
p1 + facet_wrap(~Phylum, 3)
dev.off()

Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance. This was applied in group-level comparisons for instance in Salonen et al. ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
```{r Calculate group divergences}

# Use these to compare microbiota divergence within each group

p.low <- subset_samples(pseq, mnd == "LOW")
p.low.d <- divergence(p.low)

p.med <- subset_samples(pseq, mnd == "MEDIUM")
p.med.d <- divergence(p.med)

p.high <- subset_samples(pseq, mnd == "HIGH")
p.high.d <- divergence(p.high)

p <- boxplot(list(Low = p.low.d, Medium = p.med.d, High = p.high.d))
plot(p)

```
 The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):

# VISUALISE MICORBIOME VARIATION
```{r convert to compositional data}

# Convert to compositional data
pseq.rel <- microbiome::transform(pseq, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

NMDS/Bray
Bray-Curtis dissimilarity does not consider phylogenetic relationships between OTUs. There are several distance methods and a list can be obtained by typying ?distanceMethodList in the console pane.
```{r}

pseq.ord <- ordinate(pseq.rel, "NMDS", "bray")

p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
ggsave("NMDSt.png")

p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
ggsave("NMDS_facet.png")

MDS ordination on Bray-Curtis distance
Performs principal coordinate analysis (also called principle coordinate decomposition, multidimensional scaling (MDS), or classical scaling) of a distance matrix (Gower 1966), including two correction methods for negative eigenvalues. See pcoa for further details.
```{r}

pseq.ord <- ordinate(pseq.rel, "MDS", "bray")

p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
ggsave("MDS.png")

p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
ggsave("MDS_facet.png")

PCoA for compositional data with Bray-Curtis distances
```{r}

p <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
p2 <- p + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b")

p1 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
print(p1)

```{r PCoA Bray}

# Pick core taxa
pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)

# PCoA for compositional data with Bray-Curtis distances
png("PCoA_bray.png")
p <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
                      method = "PCoA", distance = "bray") +
       labs(title = paste("PCoA / Compositional / Bray-Curtis"))
print(p)
dev.off()

t-SNE is a popular new ordination technique
, title='t-SNE (Bray)'
```{r t-SNE}

tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
  color = 'mnd')
tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
ggsave("t_sne.png")

## PERMANOVA significance test for group-level differences
Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition

```{r Redefine data}

# Convert to compositional data
pseqLvH <- merge_phyloseq(p.low,p.high)

pseq.rel <- microbiome::transform(pseqLvH, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

```{r PERMANOVA}

# samples x species as input
library(vegan)
permanova <- adonis(t(otu) ~ mnd,
               data = meta, permutations=999, method = "bray")

# P-value
print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])

Check that variance homogeneity assumptions hold (to ensure the reliability of the results):Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
```{r ANOVA}

dist <- vegdist(t(otu))
anov <- anova(betadisper(dist, meta$mnd))
summary(anov)
tidy(anov)

Show coefficients for the top taxa separating the groups
```{r top taxa L vs H}

coef <- coefficients(permanova)["mnd1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
ggsave("LvH.png")

```{r HvL}
pseqHvL <- merge_phyloseq(p.high,p.low)

pseq.rel <- microbiome::transform(pseqHvL, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

library(vegan)
permanova <- adonis(t(otu) ~ mnd,
               data = meta, permutations=99, method = "bray")

# P-value
print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])

dist <- vegdist(t(otu))
anov <- anova(betadisper(dist, meta$mnd))
summary(anov)
tidy(anov)

coef <- coefficients(permanova)["mnd1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")

```{r L vs M}

pseqLvM <- merge_phyloseq(p.low,p.med)

pseq.rel <- microbiome::transform(pseqLvM, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

library(vegan)
permanova <- adonis(t(otu) ~ mnd,
               data = meta, permutations=999, method = "bray")

# P-value
print(as.data.frame(permanova$aov.tab)["mnd", "Pr(>F)"])

dist <- vegdist(t(otu))
anov <- anova(betadisper(dist, meta$mnd))
summary(anov)
tidy(anov)

coef <- coefficients(permanova)["mnd1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus")
ggsave("LvM.png")

## MICROBIOME COMPOSITION

```{r Phylum #}

# Show available ranks in the dataset
rank_names(pseq)

# Create table, number of features for each phyla
table(tax_table(pseq)[, "Phylum"], exclude = NULL)

```
This shows a few phyla for which only one feature was observed. Those may be worth filtering.

```{r Total Phylum abundnace bar}

png("Phylum_abundance.png")
p <- plot_bar(pseq,"Phylum") + theme(axis.text.x = element_text(face="italic"))
print(p)
dev.off()

```{r Phylum prevalence table}

prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
prev.otu + theme(legend.position="none")
ggsave("phylum_prevalence.png")

tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
tab + theme(legend.position="none")
ggsave("Phylum_prevalence_detection5.png")

```{r Relative abundance}

pseq.ph <- aggregate_taxa(pseq.new, "Phylum")

pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
theme(legend.position = "bottom") + theme_bw() +
theme(axis.text.x=element_blank()) +
c + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("Phylum"))

print(plot.comp.rel)

ggsave("phylum_relative_abundance11.png")

plot.comp.rel + theme(legend.position="none")
ggsave("phylum_relative_abundance12.png")

guide_italics
ggtitle("Relative abundance") +  
```{r}

clos.ph <- aggregate_taxa(clostridium, "Phylum")

clos.ph.rel <- microbiome::transform(clos.ph, "compositional")

plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
  theme(legend.position = "bottom", ) + theme_bw() + 
  theme(axis.text.x=element_blank()) + 
  theme(legend.title = element_text(size=18), legend.text = element_text( face="italic")) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend(title="Phylum"))

print(plot.comp.rel)

ggsave("phylum_relative_abundance_clos.png")

plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none") + guide_italics
ggsave("phylum_relative_abundance_clos2.png")

```{r Phylum Bar Plot}

levels(pseq.meta$mnd)

pseq.ph <- aggregate_taxa(pseq, "Phylum")

pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))

plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
ggsave("phylum_abundance_bar.png")

```{r}

top5P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:5]

top5P = subset_taxa(pseq, Genus %in% names(top5P.names))

pseq.ph.rel <- microbiome::transform(top5P, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))

plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")

```{r}

top10P.names = sort(tapply(taxa_sums(pseq), tax_table(pseq)[, "Genus"], sum), TRUE)[1:10]

top10P = subset_taxa(pseq, Genus %in% names(top10P.names))

pseq.ph.rel <- microbiome::transform(top10P, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum"))

plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma")

```{r pseq.new bar plot}

p.new <- aggregate_taxa(pseq.new, "Phylum")

p.new.rel <- microbiome::transform(p.new, "compositional")

plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))

plot.p.new.rel + scale_fill_brewer(palette = "Paired") 
ggsave("pnew_abundance_bar.png")

```{r CLostridium Bar Plot}

clos <- aggregate_taxa(clostridium, "Phylum")

pseq.cl.rel <- microbiome::transform(clos, "compositional")

plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))

plot.cl.rel + scale_fill_brewer(palette = "Paired")
ggsave("cl_abundance_bar.png")

```{r Clostridium genus bar plots}

pseq.cl.rel <- microbiome::transform(clostridium, "compositional")

plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")

plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")

```{r Top Phylum Heatmap}

pseq.ph <- aggregate_taxa(pseq, "Phylum")

pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample", plot.type = "heatmap")

plot(plot.comp.rel+ coord_flip()) + theme(axis.text.x=element_blank())
ggsave("phylum_histogram.png")

```{r p.new heatmap}

pseq.n.ph <- aggregate_taxa(pseq.new, "Phylum")

pseq.n.ph.rel <- microbiome::transform(pseq.n.ph, "compositional")

plot.n.comp.rel <- plot_composition(pseq.n.ph.rel, x.label = "sample", plot.type = "heatmap")

p <- plot.n.comp.rel + coord_flip()
p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
ggsave("pnew_histogram.png")

```{r clostridium heatmap}

pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")

pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")

plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")

p <- plot.c.comp.rel + coord_flip()
p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
ggsave("clos_histogram.png")

```{r Network}

png("network_bray_phylum.png")
jg = make_network(pseq, "taxa", "bray", 0.3)
plot_network(jg, pseq, "taxa", color = "Phylum", line_weight = 0.4, label = NULL)
dev.off()

## Community composition plotting
Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results. These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted. Interpretation of these plots should always be subject to subsequent statistical analysis.
```{r community-composition-plots}
# Create a data table for ggplot
# agglomerate at phylum level
pseq_phy <- tax_glom(pseq.new, taxrank = "Phylum")

# Transform to rel. abundance (or use ps0.ra)
pseq_phyl <- transform_sample_counts(pseq_phy, function(x) {x/sum(x)})

# Melt to long format for easy ggploting
pseq_phylum <- psmelt(pseq_phyl)

     # filter(Abundance > 0.01)                             # Filter out low abundance taxa

# Plot - Phylum
p.ra.phylum <- ggplot(pseq_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(title = "Abundant Phylum (> 1%)")

p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
plot(p)
ggsave("phylum_abun_mnd.png")

```{r}

clos <- aggregate_taxa(clostridium, "Phylum")

# agglomerate at phylum level
clos_phy <- tax_glom(clos, taxrank = "Phylum")

# Transform to rel. abundance (or use ps0.ra)
clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})

# Melt to long format for easy ggploting
clos_phylum <- psmelt(clos_phyl)

# Plot - Phylum
p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(title = "Abundant Phylum of Clostridium(> 1%)")

p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
plot(p)
ggsave("phylum_abun_mnd_clos.png")

```{r}

get_taxa_unique(pseq, "Phylum")

```{r}

get_taxa_unique(pseq, "Genus")

```{r}

get_taxa_unique(clostridium, "Genus")

```{r}

df <- psmelt(pseq)
kable(head(df))

```{r}

head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])
